---
title: "Analyzing Spotify Top Songs 2000-2019 to help increase youtube views for aspiring instrumental cover youtubers Using R"
date: "12/12/2023"
output: rmarkdown::github_document
---

# Project Inspiration  

I have been learning the guitar for a decade and it has always been a passion of mine to share my own guitar cover videos of songs/pieces that I like. I was always in awe to discover the talents on youtube and even found it quite astonishing that some youtubers could gain a lot of views from youtube videos by just playing instruments. This showed people really appreciated true talet. Therefore, I am working on this project to try to analyze what upcoming music will be popular to help musicians to pick songs to cover to increase their popularity. To make data-driven recommendations, I set and address some questions using data analysis process: ask, prepare, process, analyze, share and act.  

# _ASK_

### Scenario

By learning from past data of listener's preference on Spotify, my task is to analyze data on Spotify and extracting insights from popular music to inform youtubers about decisions on instrumental covers. I have come up with the following questions for my analysis:

1. What are the general characteristics of popular song hits ?
2. Which song key (major/minor) tend to lead to higher popularity ?
3. What key do popular songs tend to be in ?

# _PREPARE_
The dataset about top hits from 2000-2019 is pulled from Kaggle. The details are as follows:

**Name:** Top Hits Spotify from 2000-2019\
**Credit:** MARK KOVERHA\
**Source:** Spotify\
**Link:** https://www.kaggle.com/datasets/paradisejoy/top-hits-spotify-from-20002019/data \
**Format:** CSV file, 18 columns, 2000 rows\
**Metadata:**\
- artist, song, duration_ms, explicit, year tells us some basic information of the song.\
- popularity: higher value refers to higher popularity\
- genre: genre of the song\
- danceability, energy, key, loudness,mode, speechiness, acousticness, instrumentalness, liveness, valence, tempo: music characterics of the song\

# _PROCESS_ 

```{r load_packages}
# load_packages
library(tidyverse)
library(dplyr)
library(ggplot2)
```
```{r import data}
spotify_songs <- read.csv("songs_normalize.csv")
tibble(spotify_songs)
```
```{r clean data}
#check if there is any duplicated data
spotify_songs_unique <- spotify_songs %>% distinct()
x <- nrow(spotify_songs) - nrow(spotify_songs_unique)
paste(x, "rows of data were duplicated and were removed from the dataset")
#check if there is any rows with missing information
spotify_songs_clean <- spotify_songs_unique[!apply(is.na(spotify_songs_unique) | spotify_songs_unique == "", 1, all),]
empty <- nrow(spotify_songs_unique) - nrow(spotify_songs_clean)
paste(empty, "rows of data were empty or stated not applicable and were removed from the dataset")
write.csv(spotify_songs_clean, file='spotify_cleaned.csv')
```
# Analyze
First of all, the mean, min value, max value and standard deviation were found for the attributes to learn about the general characterics of the top hits in this dataset. The general characterictics can give a hint of the metrics of an average popular song.
```{r Summary of statistics}
data <- read.csv("spotify_cleaned.csv")
#popularity
popularity_min <- min(data$popularity)
popularity_max <- max(data$popularity)
popularity_mean <- mean(data$popularity)
popularity_sd <- sd(data$popularity)
#danceability
danceability_min <- min(data$danceability)
danceability_max <- max(data$danceability)
danceability_mean <- mean(data$danceability)
danceability_sd <- sd(data$danceability)
#energy
energy_min <- min(data$energy)
energy_max <- max(data$energy)
energy_mean <- mean(data$energy)
energy_sd <- sd(data$energy)
#key
key_min <- min(data$key)
key_max <- max(data$key)
key_mean <- mean(data$key)
key_sd <- sd(data$key)
#loudness
loudness_min <- min(data$loudness)
loudness_max <- max(data$loudness)
loudness_mean <- mean(data$loudness)
loudness_sd <- sd(data$loudness)
#mode
mode_min <- min(data$mode)
mode_max <- max(data$mode)
mode_mean <- mean(data$mode)
mode_sd <- sd(data$mode)
#speechiness
speechiness_min <- min(data$speechiness)
speechiness_max <- max(data$speechiness)
speechiness_mean <- mean(data$speechiness)
speechiness_sd <- sd(data$speechiness)
#acoustiness
acousticness_min <- min(data$acousticness)
acousticness_max <- max(data$acousticness)
acousticness_mean <- mean(data$acousticness)
acousticness_sd <- sd(data$acousticness)
#instrumentalness
instrumentalness_min <- min(data$instrumentalness)
instrumentalness_max <- max(data$instrumentalness)
instrumentalness_mean <- mean(data$instrumentalness)
instrumentalness_sd <- sd(data$instrumentalness)
#liveness
liveness_min <- min(data$liveness)
liveness_max <- max(data$liveness)
liveness_mean <- mean(data$liveness)
liveness_sd <- sd(data$liveness)
#valence
valence_min <- min(data$valence)
valence_max <- max(data$valence)
valence_mean <- mean(data$valence)
valence_sd <- sd(data$valence)
#tempo
tempo_min <- min(data$tempo)
tempo_max <- max(data$tempo)
tempo_mean <- mean(data$tempo)
tempo_sd <- sd(data$tempo)
#export table
summary_table <-data.frame(
    MIN=c(popularity_min, danceability_min, energy_min,key_min, loudness_min,mode_min,speechiness_min,acousticness_min,instrumentalness_min,liveness_min,valence_min, tempo_min),
    MAX=c(popularity_max,danceability_max,energy_max,key_max,loudness_max,mode_max,speechiness_max,acousticness_max,instrumentalness_max, liveness_max, valence_max, tempo_max ),
    MEAN=c(popularity_mean,danceability_mean,energy_mean,key_mean,loudness_mean, mode_mean,speechiness_mean, acousticness_mean, instrumentalness_mean,liveness_mean, valence_mean,tempo_mean),
    SD=c(popularity_sd,danceability_sd,energy_sd,key_sd, loudness_sd,mode_sd,speechiness_sd,acousticness_sd,instrumentalness_sd, liveness_sd,valence_sd,tempo_sd))
    rownames(summary_table)<-c("popularity","danceability","energy", "key", "loundess","mode","speechiness","acousticness","instrumentalness","liveness","valence","tempo")
print(summary_table)
write.csv(summary_table,file="summary_table.csv")
```
Now, we explore how different music characteristics influences popularity. Here, we only look at songs that have high popularity. For songs to have high popularity, they have to reach higher than 75.
# Songs of high popularity
```{r high popularity}
music_characteristics_analysis <- read.csv("spotify_cleaned.csv")
high_popularity <- music_characteristics_analysis %>%  filter(popularity>=75)
total_no_of_songs <- nrow(high_popularity)
paste("Number of songs with high popularity: ",total_no_of_songs)

```
# Major/ Minor key
```{r music characteristics and popularity}
ggplot(data=high_popularity)+
    geom_bar(mapping=aes(x=mode))
major_songs <- music_characteristics_analysis %>%  filter(popularity>=75 & mode ==1)
minor_songs <- music_characteristics_analysis %>%  filter(popularity>=75 & mode == 0)
no_of_major_songs <- nrow(major_songs)
no_of_minor_songs <- nrow(minor_songs)
pie_chart_data <-c(no_of_major_songs, no_of_minor_songs)
pie_chart_labels<-c("Major","Minor")
piepercent <- round(100* pie_chart_data/sum(pie_chart_data),1)
pie(pie_chart_data,labels = piepercent, main="Major and Minor key pie chart")
legend("topright",c("Major","Minor"), cex=0.5,fill=rainbow(length(pie_chart_data)))
```
\
_Observation:_ \
From the pie chart, it was shown that songs in major key are more likely to be more popular than minor key from this dataset.

# Key
```{r key}
ggplot(data=high_popularity)+
    geom_bar(mapping=aes(x=key))
```

_Observation:_ \
From the pie chart, it was shown that songs in key 1 (C#) are the most likely to be popular while songs in key 3 (D#) are the least likely to be popular.


# _SHARE_ 
From the above observations, I have made some findings about the popular sings across 2000-2019. Generally, popular songs differ a lot in terms of Spotify metrics. For songs to be deemed to have high popularity in Spotify, they tend to be in _C#_ and _major_ key. This also gave some information about how music cover songs can be arranged in to suit the audience's taste more so they can gain more views on Spotify. 

# _ACT_
With this working hypothesis, it will be good to use these information to examine the data this year on Spotify to see if the top hit songs still follows this success formula. If this formula works, I believe aspiring muscician can take this as a source of inspiration to help build their fame on Spotify. 